Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2024-08-13 go3-author.tex
final go3 before Justin

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR) (3, 4),
exhibited as dataset-dependent FDR control. The FDR problem is connected
to the double-filtering approach that is often used, but which is known
not to have appropriate FDR control (5, 6). The second issue is poor
performance when analyzing asymmetric data; data that arises because of
a bias in the direction of change. This type of data frequently arises
in in-vitro selection experiments (SELEX), transcriptome analysis, and
microbiome analysis (7). The third issue is that these problems become
more pronounced as more samples are collected; that is, more information
results in a worsening of the accuracy of the analysis (9). The final
problem is that these data are relative data, i.e.~compositional, and
substantial effort has been put into removing this constraint.

The first three problems were recently shown by Nixon and Silverman (8)
to be a result of a mismatch between the underlying size or scale of the
system and the assumptions of the normalizations used for the analysis
of HTS. Biological variation in scale often represents an important
unmeasured confounder in HTS analyses (10). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (11), and
this dramatically skews transcriptome analysis (12). In addition,
wild-type and mutant strains of cell lines, yeast or bacteria have
different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(13--15). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (16), and the
composition between these states is dramatically different (17, 18).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale.

Recently, Nixon et al. (8) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale. These implicit assumptions are often
difficult to interpret, and different normalizations provide different
outputs when applied to the same dataset (3, 19--21). Intuitively,
normalizations in widespread use assume that either all samples have the
same scale, e.g.~proportions, rarefaction (22), RPKM (23, 24), etc; or
that a subset of features in one sample can be chosen as a reference to
which the others are scaled e.g.~the TMM,(25), or LVHA (7); or that
different sub-parts of each sample maintain a constant scale across
samples e.g.~the RLE (26); or that the geometric mean of the parts is
appropriate e.g the CLR (27) and its derivatives.

The original ALDEx2 (28) model made a strict assumption about scale
through the CLR normalization (8). While this assumption could be useful
in many cases it could never be exactly true, and others have shown that
it is not always the best option (29). Nixon et al. (8) showed that
better scale assumptions resulted in more reproducible data analysis
including better control of both false positive and false negative
results. In essence ALDEx2 has been modified to explicitly model the
scale over a range of reasonable normalization parameters. Here, we
briefly introduce these modifications and then show how scale
uncertainty can greatly improve modeling in transcriptome and
meta-transcriptome datasets to provide more robust and reproducible
results.

@@

All the normalizations in current use are data transformations that can
be stated as ratios of the form
\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
denominator is determined by some function of the observation and the
\(\hat{{\mathbf{W}}}\) notation indicates that the normalization is
attempting to provide an estimate of the true value. The technical
variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (28) is a general purpose toolbox for Bayesian
modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
form as a tool for estimating the magnitude and statistical significance
of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
expression analysis), but note that it can be used to fit more complex
linear models. At a high-level, ALDEx2 involves three steps: 1) a
Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
given observations \(\mathbf{Y}\); 2) a normalization is used to
estimate \(\mathbf{\hat{W}}\) given the estimate
\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
are used to estimate LFC as part of differential abundance (or
expression) analyses. For more details on ALDEx2 see (8, 28).

By default, ALDEx2 uses the CLR normalization which can be written as
\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
where \(G_{n}\) denotes the geometric mean of the vector
\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
normalization equates to an implicit assumption that
\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
this assumption introduces bias into the LFC estimate that is not
accounted for when estimating uncertainty (e.g., confidence intervals or
credible sets). In fact, Nixon et al. (8) showed that the only way in
which the ALDEx2 model, or any normalization-based model, could ever be
calibrated (e.g., control Type-I Error rates) was if this assumption was
exactly true. In support of this assertion it is known that false
discovery rates vary widely by analysis, dataset, and normalization
method (30, 31). To fix this, Nixon et al. (8) showed that models should
incorporate potential error in the assumptions implied by
normalizations.

@@

It is standard practice in many fields of HTS, but especially
transcriptomics, to use the dual cutoff approach graphically exemplified
by volcano plots (32, 33) because not all statistically significant
differences are biologically relevant. Paraphrasing this we can say that
in in some types of datasets more samples leads to a majority of
features being statistically significant. A detailed reasoning for this
using modelled data is explained in two reports by Nixon et al. (8, 9),
which shows that scaled models completely address this analytic issue.

We use the data of Gierliński et al. (34) who conducted a highly
replicated yeast transcriptome experiment comparing a wild-type strain
with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
argue that a dual cutoff approach is appropriate to limit the number of
significant parts and for purposes of reproducibility (33). However, in
this benchmarking study, the number of significantly different
transcripts varied between 65\% to \textgreater80\% of all transcripts
depending on the tool; in essence the desired behavior was that almost
all transcripts were significantly different. The guidance on
reproducibility runs counter to standard statistical practice where
power is intrinsically linked to sample size and very large sample sizes
are indeed desired (35). Furthermore, the dual-cutoff approach is known
not to provide appropriate FDR control (5, 6). Schurch et al. (33)
further suggested that different tools might be better in some
conditions or datasets than others because each tool has different
intrinsic statistical power and Type 1 and Type 2 errors. Through the
lens of scale uncertainty, the behaviour of the tools in this study show
unacknowledged bias; false confidence in the precision of the estimate
as sample size increases because o mis-specification in the scale of the
data. This certainty is driven by the assumptions of the tool not the
actual experiment being investigated (9).

Using either DESeq2 or ALDEx2, a majority of transcripts are
statistically significantly different between groups with a
Benjamini-Hochberg (36) false discovery rate (FDR) of 0.05; i.e.~4636
(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
large numbers of significant transcripts seems biologically unrealistic
and furthermore breaks the necessary assumption made by the
normalization methods that the majority of the features must be
invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
that the choice of normalization plays a role in which results are
returned as significant and that some if not the majority are driven by
technical differences in the analysis (20, 21, 33, 37, 38).

@@

The effect plots (39) in Figure 1A (ALDEx2) and Supplementary Figure 1
(DESeq2) shows that the majority of significant transcripts (red,
orange) have negligible differences between groups and very low
dispersion and we suggest that this is driven by the experimental design
(33). Scale uncertainty can be incorporated using the \texttt{gamma}
parameter that controls the amount uncertainty added to the CLR mean
assumption when we call either \texttt{aldex()}, or
\texttt{aldex.clr()}. Figure 1B shows that setting \(\gamma=0.5\)
results in far fewer transcripts being significant (205) and we observe
that the minimum dispersion increases from 0.12 (unscaled) to 0.67
(scaled). The Volcano plots in Figure 1 C and D show a similar story.
Here we can see that adding scale increases the minimum FDR value and
increases the concordance between the FDR value and the difference
between groups (compare panels C and D).

@@

As shown by the effect plot in Figure 1 the root cause of the many
statistically significant positive transcripts is the very large number
of transcripts with negligible variance. This phenomenon is not unique
to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon
occurs in DESeq2 (and presumably other methods although the relevant
parameters are not exposed). This issue is not unique to this dataset
and it is common practice to use a dual-cutoff by choosing transcripts
based on a thresholds for both corrected p-values and fold-changes (33)
(here set at \(\pm 2^{1.4}\) for the latter), although considerable
variation in cutoff values is observed. These limits are shown by the
dashed grey lines. Here, applying a dual-cutoff using a heuristic of at
least a \(2^{1.4}\) fold change reduces the number of significant
outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with
that observed with that found by ALDEx2 with \(\gamma = 0.5\) which
identifies 205. Indeed, Supplementary Figure 2, shows that even adding a
very small amount of scale \(\gamma = 0.1\) reduces the number of
significant transcripts by more than half and the
\texttt{aldex.scaleSim()} function can be used to identify those
transcripts that are significant only because of an absence of scale.
These results begs the question: why bother with significance tests at
all if all transcripts with \(> 1.4\)-fold expression change are
statistically significant?

The effect on dispersion with increasing amounts of scale are shown in
Figure 2A, where we can see that the dispersion increases as scale is
added. Note that the dispersion in the unscaled data in Figure 2A
reaches a minimum near the mid-point of the distribution, and also does
so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
This shows more clearly that dispersion of many transcripts is almost
negligible in the absence of scale and makes the counter-intuitive
suggestion that the variance in expression of the majority of genes with
moderate expression is more predictable than highly-expressed genes or
of housekeeping genes (40). This is at odds with the known biology of
cells where single cell counting of highly-expressed transcripts shows
that they have little intrinsic variation (41).

@@

Dos Santos et al. (42) used a vaginal metatranscriptome dataset to
compare the gene expression in bacteria collected from healthy (H) and
bacterial vaginosis (BV) affected women. In this environment, both the
relative abundance of species between groups and the gene expression
level within a species is different (43). Additionally, prior research
suggests that the total number of bacteria is about 10 times more in the
BV than in the H condition (16). Thus, this is an extremely challenging
environment in which to determine differential abundance as there are
both compositional and scale changes between conditions. The usual
method to analyze vaginal metratranscriptome data is on a taxon-by-taxon
basis (43--45) because the scale confounding can be ignored. Attempts at
system-wide analysis show that many housekeeping functions are returned
as differentially abundant between groups; a result likely due to a
disconnect between the scale assumptions of the normalization used (7).

@@

Figure 3A shows an effect plot of the data where reads are grouped by
function, corresponding to grouping homologous sequences regardless of
the organism of origin. Each point represents one of 3728 KEGG functions
(46). There are many more functions represented in the BV group (bottom)
than in the healthy group (top). This is because the
\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
reduced genome content relative to the anaerobic organisms that dominate
in BV, because there is a greater diversity of organisms in BV than in H
samples, and because the BV condition has at least an order of magnitude
more bacteria than does the H condition.

There are 101 functions with low dispersion that appear to be share by
both groups (boxed area in Figure 3A, and colored in cyan). Inspection
shows that these largely correspond to core metabolic functions such as
transcription, translation, ribosomal proteins, glycolysis, replication,
chaperones, etc (Supplementary file housekeeping.txt). The transcripts
of many of these are commonly used as invariant reference sequences (40)
and so would not be expected to contribute to differences in ecosystem
behaviour and should be centred on 0 difference. These should not be
scored as among the most differentially abundant. The major group of
these housekeeping functions is located off the line of no difference
(being approximately located at +1.5). While changes in the abundance of
housekeeping functions is a useful proxy for relative abundance of
species in the environment, they tell us nothing about the functional
capacity of the two groups as these are common to every organism. Of
more interest is determining the functions that are different between
groups because these are unique or over-expressed in one group relative
to the other.

@@

We desire a scale model that approximately centres the housekeeping
functions. We anticipate that housekeeping functions should be nearly
invariant and so an appropriate scale in this dataset is likely closer
to 0 than the naive estimate. While a user-defined scale model can be
quite flexible with relative scales that are distinct for each group (or
even each sample) along with their uncertainties, here we focus on using
the \texttt{aldex.makeScaleMatrix()} function. This function uses a
logNormal distribution to build a scale matrix given a user-specified
mean difference between groups and uncertainty level. Applying a
per-group relative differential scale of 0.15 moves the housekeeping
functions to the midline of no difference (Figure 3C), and applying a
gamma of 0.5 provides the same dispersion as in panel B). Note that now
a significant number of functions are differentially up in BV that were
formerly classed as not different without the full scale model (orange),
or when only a default scale was applied. Inspection of the functions
shows that these are largely missing from the \emph{Lactobacillus}
species and so should actually be captured as differentially abundant in
the BV group. Thus, applying a differential scale allows us to
distinguish between both false positives (housekeeping functions in
cyan, and others in blue) and false negatives (orange functions) even in
a very difficult to analyze dataset. The remarkable improvements in
biological interpretation afforded by this full scale model, and the
transferrability of it between sample cohorts of the same condition is
outlined elsewhere (42). We suggest that the default scale model is
sufficient when the data are approximately centred. However, an informed
model is more appropriate with datasets are not well centred or when the
investigator has prior information about the underlying biology.

@@

Biological systems are both predictably variable and stochastic (41) and
current measurement methods that rely on high throughput sequencing fail
to capture all of that variation, particularly variation due to scale
(8, 9). In the absence of external information (such as spike-in probes
(12), cell counting (1), FISH (47) etc) sequencing depth normalisation
methods cannot recapture the scale information (12), and can only
normalize for the technical variation due to sequencing depth.

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as
transcriptomes(4, 19, 33, 37) and microbiomes (3, 21, 29, 30, 48).
Generically, it is observed that the actual agreement between analysis
methods can be modest, and no single method outperforms all others in
every dataset or type of experiment. That different tools appear to work
more reliably in different datasets from different sources can be
explained as different normalizations being a better fit to the scale a
particular dataset by chance.

From our perspective, the disagreement between tools can be explained by
the partially identifiable nature of the data; that is, the observed
data are consistent with multiple ways of estimating the parameters (8).
Partially identifiable data exhibit multiple pathologies, chief among
them being that that different analytic approaches can produce different
parameter estimates and that more data produces worse estimates because
the additional data increases the precision of a flawed estimate (49).
In the analysis of HTS data it is often observed that larger datasets
converge on the majority of parts being significantly different (8, 33),
and that different analytic approaches result in different parts being
chosen as significantly different (4, 33, 37).

Scale simulation is now build into ALDEx2 to address the issue of
partially identifiability, and through this lens the two main root
causes to common HTS data pathologies can be proposed. The first
contributing factor is the observed very low dispersion estimate for
many features that is a by-product of experimental design, sequencing
and normalization. In the Schurch et al. (33) dataset, the data were
from single colonies derived from a single culture. Thus, it is more
accurate to describe the 96 samples as wet-lab technical replicates
rather than independent samples. However, this type of replicate is
standard in the molecular literature, and would be expected to result in
the very low dispersion that is observed.

@@

While the actual scale of the underlying environment is inaccessible
post-sequencing, we can study the sensitivity of the results to the
choice of normalization and thus scale. To do this we incorporate scale
models (8) in lieu of normalizations through the default scale model
that is now built into ALDEx2. Scale uncertainty is incorporated using
the \texttt{gamma} parameter that controls the amount uncertainty added
to the CLR mean assumption when we call either \texttt{aldex()}, or
\texttt{aldex.clr()}. The ALDEx2 package contains a sensitivity analysis
function, \texttt{aldex.senAnalysis()}, that can be used to explore the
effect of different amounts of scale uncertainty, and an example is
shown in Supplementary Figure 2 for the yeast transcriptome dataset.
Here it is clear that even tiny amounts of scale preclude the majority
of transcripts from being considered statistically significant. In
practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
is realistic for most experimental designs, but further work is needed.

@@

In the metatranscriptome analysis, transcripts that were previously not
classed as differentially abundant were now called as significantly
different, and the housekeeping transcripts move from being
significantly different to not being identified as such. Note the
assumption that housekeeping genes should not generally be included in
differential abundance analysis is implicit in the dual
p-value:fold-change cutoff approach in widespread use. While we
acknowledge that some prior information on which housekeeping
transcripts should not be classed as differentially abundant is needed,
we suggest that this information is widely available and is already used
when performing the gold-standard quantitative PCR test of differential
abundance (50, 51). Thus, the use of this prior knowledge is not unique
to our approach.

>>>>

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR)
{[}(3);(4);Nearing:\href{mailto:2022aa;@Li}{\nolinkurl{2022aa;@Li}}:2022aa{]},
exhibited as dataset-dependent FDR control. The FDR problem is connected
to the double-filtering approach that is often used, but which is known
not to have appropriate FDR control (5, 6). The second issue is poor
performance when analyzing asymmetric data; data that arises because of
a bias in the direction of change. This type of data frequently arises
in in-vitro selection experiments (SELEX), transcriptome analysis, and
microbiome analysis (7). The third issue is that these problems become
more pronounced as more samples are collected; that is, more information
results in a worsening of the accuracy of the analysis (9). The final
problem is that these data are relative data, i.e.~compositional (10,
11), and substantial effort has been put into removing this constraint.

The first three problems were recently shown by Nixon and Silverman (8)
to be a result of a mismatch between the underlying size or scale of the
system and the assumptions of the normalizations used for the analysis
of HTS. Biological variation in scale often represents an important
unmeasured confounder in HTS analyses (12). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (13), and
this dramatically skews transcriptome analysis (14). In addition,
wild-type and mutant strains of cell lines, yeast or bacteria have
different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(15--17). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (18), and the
composition between these states is dramatically different (19, 20).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale.

Recently, Nixon et al. (8) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale. These implicit assumptions are often
difficult to interpret, and different normalizations provide different
outputs when applied to the same dataset (3, 21--24). Intuitively,
normalizations in widespread use assume that either all samples have the
same scale, e.g.~proportions, rarefaction (25), RPKM (26, 27), etc; or
that a subset of features in one sample can be chosen as a reference to
which the others are scaled e.g.~the TMM (28), or LVHA (7) or the
additive log-ratio (29); or that different sub-parts of each sample
maintain a constant scale across samples e.g.~the RLE (30); or that the
geometric mean of the parts is appropriate e.g the CLR and its
derivatives.

The original ALDEx2 (31) model unwittingly made a strict assumption
about scale through the CLR normalization (8). While this assumption
could be useful in many cases it could never be exactly true, and others
have shown that it is not always the best option (32). Nixon et al. (8)
showed that better scale assumptions resulted in more reproducible data
analysis including better control of both false positive and false
negative results. In essence ALDEx2 has been modified to explicitly
model the scale over a range of reasonable normalization parameters.
Here, we briefly introduce these modifications and then show how scale
uncertainty can greatly improve modeling in transcriptome and
meta-transcriptome datasets to provide more robust and reproducible
results.

@@

All the normalizations in current use are data transformations that can
be stated as ratios of the form
\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
denominator is determined by some function of the observation. We use
the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization
is attempting to provide an estimate of the true value. The technical
variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (31, 33) is a general purpose toolbox for Bayesian
modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
form as a tool for estimating the magnitude and statistical significance
of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
expression analysis), but note that it can be used to fit more complex
linear models. At a high-level, ALDEx2 involves three steps: 1) a
Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
given observations \(\mathbf{Y}\); 2) a normalization is used to
estimate \(\mathbf{\hat{W}}\) given the estimate
\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
are used to estimate LFC as part of differential abundance (or
expression) analyses. For more details on ALDEx2 see (8, 31).

By default, ALDEx2 uses the CLR normalization which can be written as
\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
where \(G_{n}\) denotes the geometric mean of the vector
\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
normalization equates to an implicit assumption that
\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
this assumption introduces bias into the LFC estimate that is not
accounted for when estimating uncertainty (e.g., confidence intervals or
credible sets). In fact, Nixon et al. (8) showed that the only way in
which the ALDEx2 model, or any normalization-based model, could ever be
calibrated (e.g., control Type-I Error rates) was if this assumption was
exactly true. In support of this assertion it is known that false
discovery rates vary widely by analysis, dataset, and normalization
method (34, 35). To fix this, Nixon et al. (8) showed that models should
incorporate potential error in the assumptions implied by
normalizations.

@@

It is standard practice in many fields of HTS, but especially
transcriptomics, to use the dual cutoff approach graphically exemplified
by volcano plots (36, 37) because not all statistically significant
differences are biologically relevant. Paraphrasing this we can say that
in in some types of datasets more samples leads to a majority of
features being statistically significant. A detailed reasoning for this
using modelled data is explained in two reports by Nixon et al. (8, 9),
which shows that scaled models completely address this analytic issue.

We use the data of Gierliński et al. (38) who conducted a highly
replicated yeast transcriptome experiment comparing a wild-type strain
with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
argue that a dual cutoff approach is appropriate to limit the number of
significant parts and for purposes of reproducibility (37). However, in
this benchmarking study, the number of significantly different
transcripts varied between 65\% to \textgreater80\% of all transcripts
depending on the tool; in essence the desired behavior was that almost
all transcripts were significantly different. The guidance on
reproducibility runs counter to standard statistical practice where
power is intrinsically linked to sample size and very large sample sizes
are indeed desired (39). Furthermore, the dual-cutoff approach is known
not to provide appropriate FDR control (5, 6). Schurch et al. (37)
further suggested that different tools might be better in some
conditions or datasets than others because each tool has different
intrinsic statistical power and Type 1 and Type 2 errors. Through the
lens of scale uncertainty, the behaviour of the tools in this study show
unacknowledged bias; false confidence in the precision of the estimate
as sample size increases because o mis-specification in the scale of the
data. This certainty is driven by the assumptions of the tool not the
actual experiment being investigated (9).

Using either DESeq2 or ALDEx2, a majority of transcripts are
statistically significantly different between groups with a
Benjamini-Hochberg (40) false discovery rate (FDR) of 0.05; i.e.~4636
(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
large numbers of significant transcripts seems biologically unrealistic
and furthermore breaks the necessary assumption made by the
normalization methods that the majority of the features must be
invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
that the choice of normalization plays a role in which results are
returned as significant and that some if not the majority are driven by
technical differences in the analysis (22, 23, 37, 41, 42).

@@

The effect plots (43) in Figure 1A (ALDEx2) and Supplementary Figure 1
(DESeq2) shows that the majority of significant transcripts (red,
orange) have negligible differences between groups and very low
dispersion and we suggest that this is driven by the experimental design
(37). Scale uncertainty can be incorporated using the \texttt{gamma}
parameter that controls the amount uncertainty added to the CLR mean
assumption when we call either \texttt{aldex()}, or
\texttt{aldex.clr()}. Figure 1B shows that setting \(\gamma=0.5\)
results in far fewer transcripts being significant (205) and we observe
that the minimum dispersion increases from 0.12 (unscaled) to 0.67
(scaled). The Volcano plots in Figure 1 C and D show a similar story.
Here we can see that adding scale increases the minimum FDR value and
increases the concordance between the FDR value and the difference
between groups (compare panels C and D).

@@

As shown by the effect plot in Figure 1 the root cause of the many
statistically significant positive transcripts is the very large number
of transcripts with negligible variance. This phenomenon is not unique
to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon
occurs in DESeq2 (and presumably other methods although the relevant
parameters are not exposed). This issue is not unique to this dataset
and it is common practice to use a dual-cutoff by choosing transcripts
based on a thresholds for both corrected p-values and fold-changes (37)
(here set at \(\pm 2^{1.4}\) for the latter), although considerable
variation in cutoff values is observed. These limits are shown by the
dashed grey lines. Here, applying a dual-cutoff using a heuristic of at
least a \(2^{1.4}\) fold change reduces the number of significant
outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with
that observed with that found by ALDEx2 with \(\gamma = 0.5\) which
identifies 205. Indeed, Supplementary Figure 2, shows that even adding a
very small amount of scale \(\gamma = 0.1\) reduces the number of
significant transcripts by more than half and the
\texttt{aldex.scaleSim()} function can be used to identify those
transcripts that are significant only because of an absence of scale.
These results begs the question: why bother with significance tests at
all if all transcripts with \(> 1.4\)-fold expression change are
statistically significant?

The effect on dispersion with increasing amounts of scale are shown in
Figure 2A, where we can see that the dispersion increases as scale is
added. Note that the dispersion in the unscaled data in Figure 2A
reaches a minimum near the mid-point of the distribution, and also does
so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
This shows more clearly that dispersion of many transcripts is almost
negligible in the absence of scale and makes the counter-intuitive
suggestion that the variance in expression of the majority of genes with
moderate expression is more predictable than highly-expressed genes or
of housekeeping genes (44). This is at odds with the known biology of
cells where single cell counting of highly-expressed transcripts shows
that they have little intrinsic variation (45).

@@

Dos Santos et al. (46) used a vaginal metatranscriptome dataset to
compare the gene expression in bacteria collected from healthy (H) and
bacterial vaginosis (BV) affected women. In this environment, both the
relative abundance of species between groups and the gene expression
level within a species is different (47). Additionally, prior research
suggests that the total number of bacteria is about 10 times more in the
BV than in the H condition (18). Thus, this is an extremely challenging
environment in which to determine differential abundance as there are
both compositional and scale changes between conditions. The usual
method to analyze vaginal metratranscriptome data is on a taxon-by-taxon
basis (47--49) because the scale confounding can be ignored. Attempts at
system-wide analysis show that many housekeeping functions are returned
as differentially abundant between groups; a result likely due to a
disconnect between the scale assumptions of the normalization used (7).

@@

Figure 3A shows an effect plot of the data where reads are grouped by
function, corresponding to grouping homologous sequences regardless of
the organism of origin. Each point represents one of 3728 KEGG functions
(50). There are many more functions represented in the BV group (bottom)
than in the healthy group (top). This is because the
\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
reduced genome content relative to the anaerobic organisms that dominate
in BV, because there is a greater diversity of organisms in BV than in H
samples, and because the BV condition has at least an order of magnitude
more bacteria than does the H condition.

There are 101 functions with low dispersion that appear to be share by
both groups (boxed area in Figure 3A, and colored in cyan). Inspection
shows that these largely correspond to core metabolic functions such as
transcription, translation, ribosomal proteins, glycolysis, replication,
chaperones, etc (Supplementary file housekeeping.txt). The transcripts
of many of these are commonly used as invariant reference sequences (44)
and so would not be expected to contribute to differences in ecosystem
behaviour and should be centred on 0 difference. These should not be
scored as among the most differentially abundant. The major group of
these housekeeping functions is located off the line of no difference
(being approximately located at +1.5). While changes in the abundance of
housekeeping functions is a useful proxy for relative abundance of
species in the environment, they tell us nothing about the functional
capacity of the two groups as these are common to every organism. Of
more interest is determining the functions that are different between
groups because these are unique or over-expressed in one group relative
to the other.

@@

We desire a scale model that approximately centres the housekeeping
functions. We anticipate that housekeeping functions should be nearly
invariant and so an appropriate scale in this dataset is likely closer
to 0 than the naive estimate. While a user-defined scale model can be
quite flexible with relative scales that are distinct for each group (or
even each sample) along with their uncertainties, here we focus on using
the \texttt{aldex.makeScaleMatrix()} function. This function uses a
logNormal distribution to build a scale matrix given a user-specified
mean difference between groups and uncertainty level. Applying a
per-group relative differential scale of 0.15 moves the housekeeping
functions to the midline of no difference (Figure 3C), and applying a
gamma of 0.5 provides the same dispersion as in panel B). Note that now
a significant number of functions are differentially up in BV that were
formerly classed as not different without the full scale model (orange),
or when only a default scale was applied. Inspection of the functions
shows that these are largely missing from the \emph{Lactobacillus}
species and so should actually be captured as differentially abundant in
the BV group. Thus, applying a differential scale allows us to
distinguish between both false positives (housekeeping functions in
cyan, and others in blue) and false negatives (orange functions) even in
a very difficult to analyze dataset. The remarkable improvements in
biological interpretation afforded by this full scale model, and the
transferrability of it between sample cohorts of the same condition is
outlined elsewhere (46). We suggest that the default scale model is
sufficient when the data are approximately centred. However, an informed
model is more appropriate with datasets are not well centred or when the
investigator has prior information about the underlying biology.

@@

Biological systems are both predictably variable and stochastic (45) and
current measurement methods that rely on high throughput sequencing fail
to capture all of that variation, particularly variation due to scale
(8, 9). In the absence of external information (such as spike-in probes
(14), cell counting (1), FISH (51) etc) sequencing depth normalisation
methods cannot recapture the scale information (14), and can only
normalize for the technical variation due to sequencing depth.

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as
transcriptomes(4, 21, 35, 37, 41) and microbiomes (3, 23, 24, 32, 34,
52). Generically, it is observed that the actual agreement between
analysis methods can be modest, and no single method outperforms all
others in every dataset or type of experiment. That different tools
appear to work more reliably in different datasets from different
sources can be explained as different normalizations being a better fit
to the scale of a particular dataset by chance.

In the analysis of HTS data it is often observed that larger datasets
converge on the majority of parts being significantly different (8, 35,
37), and that different analytic approaches result in different parts
being chosen as significantly different (4, 24, 37, 41). Li et al. (35)
conducted a permutation-based benchmarking study and found that popular
tools performed worse than simple Wilcoxon rank-sum tests in controlling
the FDR when sample sizes became large. They suggested that the presence
of outliers were one of the factors driving this observation. Brooks et
al. (53) suggested that inappropriate choice of benchmarking methods are
also a major contributing factor. From our perspective, the disagreement
between tools can be explained by the partially identifiable nature of
the data; that is, the observed data are consistent with multiple ways
of estimating the parameters (8). Partially identifiable data exhibit
multiple pathologies, chief among them being that that different
analytic approaches can produce different parameter estimates and that
more data produces worse estimates because the additional data increases
the precision of a flawed estimate (54).

Scale simulation is now build into ALDEx2 to address the issue of
partially identifiability, and through this lens the two main root
causes to common HTS data pathologies can be proposed. The first
contributing factor is the observed very low dispersion estimate for
many features that is a by-product of experimental design, sequencing
and normalization. In the Schurch et al. (37) dataset, the data were
from single colonies derived from a single culture. Thus, it is more
accurate to describe the 96 samples as wet-lab technical replicates
rather than independent samples. However, this type of replication
approach is standard in the molecular literature, and would be expected
to result in the very low dispersion that is observed.

@@

While the actual scale of the underlying environment is inaccessible
post-sequencing, we can study the sensitivity of the results to the
choice of normalization and thus scale. Varying amounts of scale
uncertainty is incorporated using the \texttt{gamma} parameter that
controls the amount uncertainty added to the CLR mean assumption when we
call either \texttt{aldex()}, or \texttt{aldex.clr()}. The ALDEx2
package contains a sensitivity analysis function,
\texttt{aldex.senAnalysis()}, that can be used to explore the effect of
different amounts of scale uncertainty, and an example is shown in
Supplementary Figure 2 for the yeast transcriptome dataset. Here it is
clear that even tiny amounts of scale preclude the majority of
transcripts from being considered statistically significant. In
practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
is realistic for most experimental designs, but further work is needed.

@@

In the metatranscriptome analysis, transcripts that were previously not
classed as differentially abundant were now called as significantly
different, and the housekeeping transcripts move from being
significantly different to not being identified as such. Note the
assumption that housekeeping genes should not generally be included in
differential abundance analysis is implicit in the dual
p-value:fold-change cutoff approach in widespread use. While we
acknowledge that some prior information on which housekeeping
transcripts should not be classed as differentially abundant is needed,
we suggest that this information is widely available and is already used
when performing the gold-standard quantitative PCR test of differential
abundance (55, 56). Thus, the use of this prior knowledge is not unique
to our approach.
